Refunctionalization of Abstracting Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
section 7

To begin, we now present the concrete syntax of a call-by-value $\lambda$-calculus language
in ANF~\cite{flanagan1993essence}.

@@

A history of the last $k$ call sites is used as a finite program contour in
$k$-call-sensitive analysis.
The history is represented as a list of expressions and embedded
in the allocated addresses.

Every time before transferring to successor states, we use function @tick@ to
fresh the time stamp and use this new time stamp for successors when allocating addresses.
The function @tick@ returns the $k$ front-most expressions given the current state and its
time history.

@@

If we instantiate $k$ to be $0$, the history degenerates to an empty list,
and we obtain a monovariant analysis, which does
not differentiate values at different call sites. In this case, the address
space collapses to the space of variable names.
Note that regarding the ambiguity in $k$-CFA\cite{Gilray:2016:ACP:2951913.2951936},
the code here actually implements call+return sensitivity.

@@

To run (analyze) a program, as same as the CESK machines,
we first use function @inject@ to construct the initial state given the program.
Note that the initial continuation store has a built-in mapping that maps continuation address
@Halt@ to an empty set of continuations.
We also provide an empty program contour as initial time.
\begin{lstlisting}
def inject(e: Expr): State =
  State(e, Map(),
        Store[BAddr, Storable](Map()),
        Store[KAddr, Cont](Map(Halt -> Set())),
        Halt, List())
\end{lstlisting}

However, comparing with the concrete CESK machine, the
@drive@ function performs collecting semantics instead of the valuation
semantics. That is, for the purpose of analyzing programs, the function @drive@
collects all the intermediate machine states as the program is abstractly executing.
The following code shows a variant of the work list algorithm to find the fixed-point
of set of states.
Function @drive@ always applies function @step@ to the head element @hd@ of
the work list @todo@ if @hd@ is unseen, then inserts the result of @step@ to
the rest of work list; and in the meantime adds @hd@ to the explored states set.
If the work list is empty, then @drive@ just returns the set of reachable states
up to now.

@@

Finally, a user could use function @analyze@ to obtain all reachable states for a given program.

@@

In this section, we describe a variant of AAM that leaves the stack to be
unbounded. That uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding
addresses of continuations in states, we intend to use a list of frames
to explicitly model the stack.
By doing so, we recover the call stack as precise as runtime's
(so called pushdown control flow analysis), but since
the stack is unbounded, the analysis is potentially not computable.
For readers who are not familiar with pushdown analysis, we have a detailed
discussion in Section \ref{pdcfarevisit}.
The reason we show it here is to establish an artifact with precise call/return match
used for the next step of transformation.

In the definition of state @State@, the continuation store disappears, and
component @konts@ becomes a list of frames. The other components keep the same.

@@

The state transition function @step@ shows below is still
nondeterministic, but the only nondeterminism happens when dereferencing the callee @f@
from function applications @App(f, ae)@.

@@

In the first case of pattern matching, we may have multiple choices of closure for callee @f@.
For each closure in the set, a new frame is constructed and pushed onto the stack.
The code for handling the second case is the same as the concrete CESK machines.

\textbf{Note on Computability.}
Unfortunately, even though, other components in states are finite,
this AAM with unbounded stack is still not computable.
Because the unbounded stack can grow to arbitrary depth which implies that the state space is possibly
infinite, therefore the analysis may not terminate for all programs if we simply enumerate reachable states.
To see this, consider a program that has two mutually recursive functions:

@@

Function @f1@ and @f2@ mutually invokes each other, and the stack will be
alternatively pushing frame @f1@ and @f2@ onto the top of current stack,
but no existing two stack components are identical in the state space.

@@

In the last section, we showed that by keeping an unabstracted stack in the state
space we can recover the precise call/return match.
From this section, we start describing the transformations step by step.
Our base machine is the AAM with unbounded stack.
In fact, as later we will transform the stack to higher-order functions
represented continuations, it does not matter what kind of AAM we start from.
Because transforming to higher-order functions forces us to sequentialize
the order of abstract evaluation, and neither need to construct a frame of list
nor to allocate continuations in the store.
We choose to start from AAM with unbounded stack just because it has a
equivalence of stack model to the final abstract definitional interpreters.

@@

On another path, here, we introduce direct-style abstracting definitional interpreter
with delimited control operators @reset@ and @shift@.

TODO: This version does not use delimited controls.

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache): Ans = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) {
    return Ans(cache.outGet(config), cache)
  }

  var new_cache = cache.outUpdate(config, cache.inGet(config))
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = aeval(f, env, store).toList.asInstanceOf[List[Clos]]
      val letvs = for (Clos(Lam(v, body), c_env) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, aeval(ae, env, store))
        val Ans(bdv, bdcache) = aeval(body, new_env, new_store, new_time, new_cache)
        new_cache = bdcache
        val evs = for (VS(vals, time, store) <- bdv) yield {
          val baddr = allocBind(x, time)
          val new_env = env + (x -> baddr)
          val new_store = store.update(baddr, vals)
          val Ans(ev, ecache) = aeval(e, new_env, new_store, time, new_cache)
          new_cache = ecache
          ev
        }
        evs.foldLeft(Set[VS]())(_ ++ _)
      }
      val letv = letvs.foldLeft(Set[VS]())(_ ++ _)
      Ans(letv, new_cache.outUpdate(config, letv))

    case ae if isAtomic(ae) =>
      val vs = Set(VS(aeval(ae, env, store), new_time, store))
      Ans(vs, new_cache.outUpdate(config, vs))
  }
}
\end{lstlisting}

>>>>

To begin, we present the concrete syntax of a call-by-value $\lambda$-calculus language
in ANF~\cite{flanagan1993essence}.

@@

In $k$-call-sensitive analysis, a history of the last $k$ call sites is
used as a finite program counter.
The history is represented as a list of expressions and embedded
in the allocated addresses.

Before transferring to successor states, we must use the @tick@ function to
refresh the timestamp, and then use this new timestamp for successors when allocating addresses.
The @tick@ function returns the $k$ front-most expressions given the current state and its
time history.

@@

If we instantiate $k$ to be $0$, the history degenerates to an empty list,
and we obtain a monovariant analysis (i.e., it does
not differentiate values at different call sites). In this case, the address
space collapses to the space of variable names.
Note that regarding the ambiguity in $k$-CFA\cite{Gilray:2016:ACP:2951913.2951936},
the code here actually implements call+return \note{call+return, call/return, etc.? Check this} sensitivity.

@@

Similar to the CESK machines, to run (analyze) a program
we first use the @inject@ function to construct the initial state given to the program.
Note that the initial continuation store has a built-in mapping that maps the continuation
address for @Halt@ to an empty set of continuations.
We also provide an empty program counter as our initial time.
\begin{lstlisting}
def inject(e: Expr): State =
  State(e, Map(),
        Store[BAddr, Storable](Map()),
        Store[KAddr, Cont](Map(Halt -> Set())),
        Halt, List())
\end{lstlisting}

However, in contrast to the concrete CESK machine, the
@drive@ function performs collecting semantics instead of the valuation
semantics. That is, for the purpose of analyzing programs, the function @drive@
collects all the intermediate machine states as the program is abstractly executing.
The following code shows a variant of the worklist algorithm to find the fixed-point
of the set of states.\note{JD: I changed all of this to worklist instead of work list in favor of the worklist algorithm. Change this back if it's wrong.}
Function @drive@ always applies function @step@ to the head element @hd@ of
the worklist @todo@ if @hd@ is unseen. It then inserts the result of @step@ to
the rest of worklist, and in the meantime adds @hd@ to the explored states set.
If the worklist is empty, @drive@ simply returns the set of reachable states
up to the current execution point.

@@

Finally, a user may invoke the @analyze@ function to obtain all reachable states for a given
program.

@@

In this section, we describe a variant of AAM that allows the stack to be
unbounded which uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding
addresses of continuations in states, we intend to use a list of frames
to explicitly model the stack.
By doing so, we recover the call stack as precise as runtimes
(so called pushdown control flow analysis), but since
the stack is unbounded, the analysis is potentially not computable.
For readers who are not familiar with pushdown analysis, we have a detailed
discussion in Section~\ref{pdcfarevisit}.
The reason we show it here is to establish an artifact with precise call/return matching
used for the next step of transformation.

In the definition of @State@, the continuation store disappears, and
component @konts@ becomes a list of frames. The other components remain unchanged.

@@

The state transition function @step@ shown below is still
nondeterministic, but the only nondeterminism happening is when dereferencing the callee @f@
from the function application @App(f, ae)@.

@@

In the first case of pattern matching, we may have multiple choices of closure for callee @f@.
For each closure in the set, a new frame is constructed and pushed onto the stack.
The code for handling the second case (atomic expressions) is the same as the concrete CESK
machines.

\textbf{Note on Computability.}
Unfortunately, even though other components in the states are finite,
this AAM with an unbounded stack is still not computable.
This is because the unbounded stack can grow to arbitrary depth which implies that the state
space is possibly infinite; the analysis may therefore not terminate for all programs if we
simply enumerate reachable states. To see this, consider a program that has two mutually
recursive functions:

@@

Function @f1@ and @f2@ mutually invoke each other, so the stack will
alternate pushing frames @f1@ and @f2@ onto the top of current stack.
However, no two existing stack components are identical in the state space.

@@

In the previous section, we show that by keeping an unabstracted stack in the state
space, we can recover the precise call/return match.
In this and following sections, we begin describing the transformations step by step.
Our base machine for now is the AAM with unbounded stack, though
as we will later transform the stack to higher-order functions
represent \note{``representing'' or ``represented by''?} continuations, it does not matter what kind of AAM we start from.
This is because transforming to higher-order functions forces us to sequentialize
the order of abstract evaluation, and neither versions of AAM requires construction of a frame of lists \note{should this be frame of lists? sounds weird}, nor to allocate continuations in the store.
Thus, our choice to start from an AAM with unbounded stack is motivated simply because it has an equivalent stack model to abstract definitional interpreters (our final target).

@@

As we identified before,

\begin{lstlisting}
\end{lstlisting}
